---
title: "Peer Assessment 1"
output: html_document
---

Peer Assessment 1
==================

## Loading and preprocessing the data

```{r}
unzip(zipfile="activity.zip")
data <- read.csv(file="activity.csv")
```

## What is mean total number of steps taken per day?
For this part, we ignore the missing values in the dataset.
```{r}
steps_by_date <-tapply(data$steps, data$date, sum)
hist(steps_by_date, 
     xlab="Number of steps",
     main="Total number of steps taken each day")
```


```{r}
mean(steps_by_date, na.rm=TRUE)
median(steps_by_date, na.rm=TRUE)
```
The mean is `r mean(steps_by_date, na.rm=TRUE)`.
The median is `r median(steps_by_date, na.rm=TRUE)`.

## What is the average daily activity pattern?
```{r}
steps_by_interval <-tapply(data$steps, 
                           data$interval, 
                           function(x) mean(x, na.rm=TRUE))
plot(steps_by_interval, type = "l", main="Average number of steps", 
     xlab="5-minute interval", ylab="Average number of steps")
```

```{r}
names(steps_by_interval[which.max(steps_by_interval)])
```
On average across all the days in the dataset, the 5-minute interval 
`r names(steps_by_interval[which.max(steps_by_interval)])` contains the maximum 
number of steps.

## Imputing missing values
```{r}
sum(is.na(data$steps))
```
The total number of missing values in the dataset is `r sum(is.na(data$steps))`.

```{r}
a<-tapply(data$steps, data$interval, function(x) sum(!is.na(x)))
b<-tapply(data$steps, data$date, function(x) sum(!is.na(x)))
# number of non NA by interval
sum(a==0)
# number of non NA by date
sum(b==0)
```
For filling in all of the missing values in the dataset, I use the mean for the 
5-minute interval because for one missing value, all the values for that day 
are missing.

I create a new dataset that is equal to the original dataset but with the 
missing data filled in.
```{r}
new_data <- data
mean_by_interval <- tapply(new_data$steps, 
                           new_data$interval, 
                           function(x) mean(x, na.rm=TRUE))

for (i in seq(1,nrow(new_data))) {
  if (is.na(new_data$steps[i])) {
    new_data$steps[i] <- mean_by_interval[as.character(new_data$interval[i])]
  }
}
```


```{r}
steps_by_date <-tapply(new_data$steps, new_data$date, sum)
hist(steps_by_date, xlab="Number of steps", 
     main="Total number of steps taken each day")
```

```{r}
mean(steps_by_date)
median(steps_by_date)
```
The mean is `r mean(steps_by_date)`.
The median is `r median(steps_by_date)`.

These values don't differ from the estimates from the first part. Imputing 
missing data impact on the estimates of the total daily number of steps: it 
adds 8 days at the average.


## Are there differences in activity patterns between weekdays and weekends?

```{r}
new_data$date <- strptime(new_data$date, format="%Y-%m-%d")
new_data$weekday <- factor(weekdays(new_data$date) %in% c("Saturday", "Sunday"),
                           labels=c("weekday", "weekend"))
```

```{r}
library(lattice)
averages<-tapply(new_data$steps, list(new_data$interval,new_data$weekday), mean)

for (i in seq(1,nrow(new_data))) {
    new_data$average[i] <- averages[as.character(new_data$interval[i]), 
                                    new_data$weekday[i]]
}
xyplot(average ~ interval | weekday, data=new_data, type="l", layout=c(1,2), 
       ylab="Average number of steps taken", 
       main="Average number of steps by day and interval")
```
